---
title: "Aviation Crash Analysis Final Notebook"
output: html_notebook
---

## Data Cleansing

# Import data and packages

```{r}

#install.packages("readr")

library(dplyr)
library(readr)
library(zoo)
library(lubridate)
library(missForest)

aero <- read.csv("Airplane_Crashes_and_Fatalities_Since_1908.csv")

```


# check data format

```{r}

str(aero)

```

We can see the following data in aero:

Date - This is the date of the crash
Time - This is the time the crash occurred
Location - The location of the crash
Operator - This is the flight operator
Type - The type of crashed aircraft
Registration - aircraft registration
cn.In - Construction or serial number/Line or fuselage number is the unique identifier bestowed upon an aircraft during its creation
Aboard - The number of passenger and crew fatalities
Fatalities - The total number of fatalities
Ground - The number of casualties not on-board the aircraft
Summary - A summary of the crash


We can see a few of the values are already in int format. We can focus on these first.

# Percentage of missing values in each column

```{r}

misscol <- colSums(is.na(aero)) / nrow(aero)
round(misscol, 3)

```

We can see that Aboard, Fatalities and Ground all have less than 1% missing data. Imputation of NA values with the mean
will be a safe bet not to BIAS

# Replace NA with the column mean using zoo and round to a whole number

```{r}

# Fatalities

aero$Fatalities <- na.aggregate(aero$Fatalities)
aero$Fatalities <- round(aero$Fatalities)

# Ground

aero$Ground <- na.aggregate(aero$Ground)
aero$Ground <- round(aero$Ground)

# Aboard

aero$Aboard <- na.aggregate(aero$Aboard)
aero$Aboard <- round(aero$Aboard)

```


Now let's create new features

```{r}

# Calculate total fatalities

aero$total_fatalities <- aero$Fatalities + aero$Ground

# Categorize severity
aero$crash_severity <- ifelse(aero$total_fatalities == 0, "Low",
                                          ifelse(aero$total_fatalities < 50, "Medium", "High"))

# verify results

head(aero)  # To verify the results
```



For the other values, let us remove missing spaces and dashes and convert these values into NA
Date and Time we can focus on later.

# Changing missing spaces and dashes into NA

```{r}

# Changing missing spaces and dashes into NA

aero$Location <- ifelse(aero$Location == "", NA, aero$Location)
aero$Operator <- ifelse(aero$Operator == "", NA, aero$Operator)
aero$Flight.. <- ifelse(aero$Flight.. == "", NA, aero$Flight..)
aero$Flight.. <- ifelse(aero$Flight.. == "-", NA, aero$Flight..)
aero$Route <- ifelse(aero$Route == "", NA, aero$Route)
aero$Type <- ifelse(aero$Type == "", NA, aero$Type)
aero$Registration <- ifelse(aero$Registration == "", NA, aero$Registration)
aero$cn.In <- ifelse(aero$cn.In == "", NA, aero$cn.In)
aero$Summary <- ifelse(aero$Summary == "", NA, aero$Summary)

aero <- aero[!is.na(aero$Summary), ] # Remove NA Summary rows

```


```{r}

head(aero$Date)

```


# Convert Date to a date type variable

```{r}

aero$Date <- as.Date(aero$Date, format = "%m/%d/%Y")

```


# Percentage of missing values in each column

```{r}

misscol <- colSums(is.na(aero)) / nrow(aero)
round(misscol, 3)

```

Date has no missing values

Summary will be useful for creating new columns.

Location, Operator and Type all have less than 1% missing values

# Replace NA in Location, Operator and Type with "Unavailable"

```{r}

# replace NA with "unavailable"

aero <- aero %>%
  dplyr::mutate(Location = replace(Location, is.na(Location), "Unavailable")) %>%
  dplyr::mutate(Type = replace(Type, is.na(Type), "Unavailable")) %>%
  dplyr::mutate(Operator = replace(Operator, is.na(Operator), "Unavailable"))



```


Let's fill in missing spaces in Time as NA convert time into a factor and take a look at the missing data again.

```{r}
# Fill in empty string or those with dashes with NA and change into a factor

aero$Time[aero$Time == ""] <- NA 
aero$Time <- ifelse(aero$Time == "-", NA, aero$Time)
aero$Time <- factor(aero$Time)

```


```{r}

# Check for missing values

misscol <- colSums(is.na(aero)) / nrow(aero)
round(misscol, 3)

```

Time has 42.1% missing data and may need to be dropped
Flight has 81% missing data and may need be dropped

Before deciding on Feature Selection I should convert the char values into categorical values and see their correlation


Time could potentially be added to date using the lubridate package, however, the data may prove useful if shrunk down it smaller Categorical times of the day.


We should first check for incorrect variables and correct them if found

```{r}

wrong_Values <- aero$Time[!grepl("^\\d{2}:\\d{2}$", aero$Time)]

#Print the non-conforming values
print(wrong_Values[!is.na(wrong_Values)])

```

We can see a few non conforming values, such as c: 1:00 12'20. Let's fix this first 

```{r}

aero$Time <- gsub("^c", "", aero$Time)  # Remove "c:" prefix
aero$Time <- gsub("'", ":", aero$Time) # Replace the apostrophe with a colon
aero$Time <- gsub("^:", "", aero$Time) # Remove leading colon
aero$Time <- trimws(aero$Time) # remove leading space
aero$Time <- sprintf("%02s",aero$Time) # add leading zero

```

We should check to see if there are any remaining incorrect values

```{r}

# Refresh non_conforming_values

wrong_Values <- aero$Time[!grepl("^\\d{2}:\\d{2}$", aero$Time)]

# We can see a few values still require fixing

unique(wrong_Values)

```


There are still incorrect values, though not many. We can fix the values manually though will need their index positions.


```{r}

# Some values aren't fixed by adding leading zeros. Let's get their indexes and fix them manually

index_of_2_00 <- which(aero$Time == "2:00")
print(index_of_2_00)

index_of_1_00 <- which(aero$Time == "1:00")
print(index_of_1_00)

index_of_1_30 <- which(aero$Time == "1:30")
print(index_of_1_30)

index_of_18_40 <- which(aero$Time == "18.40")
print(index_of_18_40)

index_of_114_20 <- which(aero$Time == "114:20")
print(index_of_114_20)

index_of_09430 <- which(aero$Time == "0943")
print(index_of_09430)

index_of_2_40 <- which(aero$Time == "2:40")
print(index_of_2_40)

index_of_9_40 <- which(aero$Time == "9:40")
print(index_of_9_40)

index_of_8_02 <- which(aero$Time == "8:02")
print(index_of_8_02)

index_of_9_30 <- which(aero$Time == "9:30")
print(index_of_9_30)
```

Now we will replace the incorrectr values at their index positions.

```{r}

# Manually correct values without leading zeroes or missing colons that for some reason won't adjust.

new_value1 <- "02:00"
aero$Time[229] <- new_value1

new_value2 <- "02:00"
aero$Time[4849] <- new_value2

new_value3 <- "01:00"
aero$Time[191] <- new_value3

new_value4 <- "01:00"
aero$Time[3585] <- new_value4

new_value5 <- "01:30"
aero$Time[712] <- new_value5

new_value6 <- "18:40"
aero$Time[1950] <- new_value6

new_value7 <- "14:20"
aero$Time[3268] <- new_value7

new_value8 <- "09:43"
aero$Time[3537] <- new_value8

new_value9 <- "02:40"
aero$Time[4299] <- new_value9

new_value10 <- "09:40"
aero$Time[4839] <- new_value10

new_value11 <- "08:02"
aero$Time[4850] <- new_value11

#new_value12 <- "09:30"
#aero$Time[5157] <- new_value12

```



# Create a new column named TimeOfDay and add Categorical data


```{r}

library(lubridate)

# Convert the format to DateTime and extract the hour

aero$Time <- as.POSIXct(aero$Time, format = "%H:%M")
aero$Hour <- hour(aero$Time)


# Take the hour and depending on its value change it into a category

aero$TimeOfDay <- cut(
  aero$Hour,
  breaks = c(0, 6, 12, 18, 24),
  labels = c("Night", "Morning", "Afternoon", "Evening"),
  include.lowest = TRUE # Includes 0:00 in Night
)

```

# Encoding TimeOfDay

```{r}

# Encode TimeOfDay 

aero$TimeOfDay <- as.factor(aero$TimeOfDay)  # Convert to a factor
encoded_data <- data.frame(TimeOfDay = as.integer(aero$TimeOfDay))

aero$TimeOfDay <- encoded_data$TimeOfDay

```

# Drop Time and Hour from the Dataset

```{r}

library(dplyr)
aero <- aero %>%
  select(-Time, -Hour)

```

TimeOfDay is still missing 42.1% of data but we'll look at creating other variables before deciding on any imputation.


# Search for most frequent words in Summary (Text Mining)

```{r}
library(tidytext)
library(dplyr)

aero <- as_tibble(aero)

# Tokenize and count words
word_counts <- aero %>%
  unnest_tokens(word, Summary) %>%  # This function splits the text into individual words.
  count(word, sort = TRUE)          # Count occurrences of each word

print(word_counts)

```


# Word PreProcessing

```{r}

library(tidyverse)
library(tidytext)
library(SnowballC)

word_counts <- word_counts %>%
  mutate(word = tolower(word)) %>% # make all words lowercase
  mutate(word = str_trim(word)) %>% # remove white space
  mutate(stem = wordStem(word)) %>% # stem words
  filter(!word %in% stop_words$word) %>% # remove English stop words
  filter(!grepl("^\\d+(,\\d+)*$", word)) %>% # remove numbers with optional commas
  filter(!grepl("^\\d+(\\.\\d+)?$", word)) %>% # remove numbers with optional decimal places
  filter(!grepl("^[^[:alpha:][:punct:]]+$", word)) # remove punctuation and any character that is not a letter
 

print(word_counts)

```


# Create new column for cause of crash

```{r}

# Create a function to map tokens to cause_of_crash values

get_cause_of_crash <- function(tokens) {
  tokens <- gsub("[\\.,]", "", tokens)
  tokens <- tolower(tokens)
  if ("fog" %in% tokens || "storm" %in% tokens || "weather" %in% tokens ||
      "poor" %in% tokens || "rain" %in% tokens || "icing" %in% tokens ||
      "thunderstorm" %in% tokens || "ice" %in% tokens || "thunderstorms" %in% tokens ||
      "storm" %in% tokens || "lightning" %in% tokens || "snow" %in% tokens ||
      "rainstorm" %in% tokens || "gust" %in% tokens || "overcast" %in% tokens ||
      "winds" %in% tokens || "snowstorm" %in% tokens || "winds" %in% tokens ||
      "volcano" %in% tokens || " hurricane" %in% tokens || "tornado" %in% tokens ||
      "meteorological" %in% tokens){
    return("Bad Weather / Natural Disaster")
  } else if ("pilot" %in% tokens || "crew" %in% tokens || "pilot's" %in% tokens ||
             "captain" %in% tokens || "pilots" %in% tokens || "decision" %in% tokens ||
             "fatigue" %in% tokens || "procedure" %in% tokens || "training" %in% tokens ||
             "operation" %in% tokens || "captain's" %in% tokens || "misjudged" %in% tokens ||
             "planning" %in% tokens || "officer" %in% tokens || "judgement" %in% tokens ||
             "experience" %in% tokens || "instructions" %in% tokens || "judgment" %in% tokens ||
             "management" %in% tokens || "crews" %in% tokens || "awareness" %in% tokens ||
             "flightcrew" %in% tokens || "flightcrew's" %in% tokens || "preparation" %in% tokens ||
             "engineer" %in% tokens || "exhaustion" %in% tokens || "operator" %in% tokens ||
             "alcohol" %in% tokens || "incapacitation" %in% tokens || "officer's" %in% tokens) {
    return("Human Error")
  } else if ("war" %in% tokens || "terrorism" %in% tokens || "political" %in% tokens || "shot" %in% tokens || 
             "hijackers" %in% tokens || "suicide" %in% tokens || "hijacker" %in% tokens || "enemy" %in% tokens ||
             "fighters" %in% tokens || "hijacked" %in% tokens || "bomb"  %in% tokens || "bombs" %in% tokens ||
             "mission" %in% tokens || "explosive" %in% tokens || "detonation" %in% tokens || "missile" %in% tokens ||
             "fighter" %in% tokens || "military" %in% tokens || "aircraft" %in% tokens && "fire" %in% tokens ||
             "shelling" %in% tokens) {
    return("Act of War / Terrorism")
  } else if ("mechanical" %in% tokens || "failure" %in% tokens || "engine" %in% tokens || "wing" %in% tokens ||
             "power" %in% tokens || "engines" %in% tokens || "instrument" %in% tokens || "nose" %in% tokens ||
             "gear" %in% tokens || "system" %in% tokens || "propeller"  %in% tokens || "burst" %in% tokens ||
             "aircraft's" %in% tokens || "cockpit"  %in% tokens || "radar" %in% tokens || "fuselage" %in% tokens ||
             "flaps" %in% tokens || "navigational" %in% tokens || "rudder"  %in% tokens || "jet" %in% tokens ||
             "separation" %in% tokens || "controls" %in% tokens || "instruments" %in% tokens || "structural" %in% tokens ||
             "wings" %in% tokens || "tank" %in% tokens || "design" %in% tokens || "door" %in% tokens || "electrical" %in% tokens ||
             "smoke" %in% tokens || "rear" %in% tokens || "malfunction" %in% tokens || "stalled" %in% tokens || "stall" %in% tokens ||
             "pressure" %in% tokens || "stabilizer" %in% tokens|| "equipment" %in% tokens || "uncontrollable" %in% tokens || 
             "altimeter" %in% tokens || "blade" %in% tokens || "feathered" %in% tokens || "uncontrolled" %in% tokens ||
             "faulty" %in% tokens || "deteriorating" %in% tokens || "device" %in% tokens || "starboard" %in% tokens ||
             "mechan" %in% tokens || "autopilot" %in% tokens || "malfunctioning" %in% tokens || "airplane's" %in% tokens ||
             "plane's" %in% tokens || "errors" %in% tokens || "contamination" %in% tokens || "tanks"%in% tokens ||
             "developed" %in% tokens || "fracture" %in% tokens || "inoperative" %in% tokens || "valve" %in% tokens ||
             "cylinder"  %in% tokens || "feather" %in% tokens || "breaking" %in% tokens || "damaged" %in% tokens ||
             "rotation"  %in% tokens || "wheel" %in% tokens || "drifted" %in% tokens || "ignition" %in% tokens || "turbine" %in% tokens ||
             "crack" %in% tokens || "propellers" %in% tokens || "blades" %in% tokens || "bolts" %in% tokens || "corrosion" %in% tokens ||
             "cracks" %in% tokens || "decompression" %in% tokens || "metal" %in% tokens || "rod" %in% tokens || "ruptured" %in% tokens ||
             "wires" %in% tokens || "broken" %in% tokens || "leak" %in% tokens || "throttle" %in% tokens || "bolt" %in% tokens || 
             "compressor" %in% tokens || "gas" %in% tokens || "generator" %in% tokens || "loose" %in% tokens || "pump" %in% tokens ||
             "switch" %in% tokens || "brake" %in% tokens || "brakes" %in% tokens || "cables" %in% tokens || "handle" %in% tokens)  {
    return("Mechanical Failure")
  } else {
    return("Other")
  }
}

# Apply the function to each row

aero$Cause_of_crash <- sapply(aero$Summary, function(summary) {
  tokens <- strsplit(summary, "\\s+")[[1]]  # Split summary into tokens
  get_cause_of_crash(tokens)
})
```


# Create a Phase of Flight column

```{r}

get_Phase_of_Flight <- function(tokens) {
  tokens <- gsub("[\\.,]", "", tokens)
  tokens <- tolower(tokens)
  if("takeoff" %in% tokens){
    return("Takeoff")
}else if("landing" %in% tokens || "land" %in% tokens)
    return("Landing")
  else{
  return("In Flight")
  }
}

aero$Phase_of_Flight <- sapply(aero$Summary, function(summary) {
  tokens <- strsplit(summary, "\\s+")[[1]]  # Split summary into tokens
  get_Phase_of_Flight(tokens)
})
```


# LDA

```{r}

# Load the required package
library(tm)
library(topicmodels)



# Create a corpus from the 'notes' column

my_corpus <- Corpus(VectorSource(aero$Summary))

# Perform tokenization

my_corpus <- tm_map(my_corpus, content_transformer(tolower))  # Convert to lowercase
my_corpus <- tm_map(my_corpus, removePunctuation)             # Remove punctuation
my_corpus <- tm_map(my_corpus, removeNumbers)                 # Remove numbers
my_corpus <- tm_map(my_corpus, removeWords, stopwords("en"))  # Remove common English stopwords
my_corpus <- tm_map(my_corpus, stripWhitespace)               # Remove extra whitespaces

# Inspect the first few tokens

inspect(my_corpus[1:5])



```

```{r}

# Create a document-term matrix
dtm <- DocumentTermMatrix(my_corpus)

# Convert the DTM to a matrix
dtm_matrix <- as.matrix(dtm)

#  convert to a data frame for easier manipulation
dtm_df <- as.data.frame(dtm_matrix)

# Assign meaningful column names (terms)
colnames(dtm_df) <- make.names(colnames(dtm_df))

# Now you have your DTM ready!

```

```{r}

# Calculate the sum of each row in `dtm`
raw.sum <- apply(dtm, 1, sum)

# Filter out rows with zero sum (if needed)
dtm <- dtm[raw.sum != 0, ]

# Specify the number of topics (e.g., 4)
num_topics <- 6

# Fit the LDA model
lda_model <- LDA(dtm, k = num_topics)

# Get the terms associated with each topic
terms_per_topic <- terms(lda_model, 10)  # Adjust the number of terms as needed

# Assign topics to your crash notes
topic_assignments <- topics(lda_model)

# Create a new data frame with the topic assignments
aero <- data.frame(aero, topic = factor(topic_assignments))



```

```{r}
str(aero)
```

```{r}


```

